### 2010-03-31 papers/icfp-2010-desc/paper_type_theory.tex
the basic type theory overhauled

Our work is deeply rooted in the realm of dependently-typed functional
languages. In particular, we have developed this technology in our
on-going work on Epigram. However, in this paper,
we strive to present our technology in the broadest setting
possible. To this end, we are going to present a type theory from the
ground up.

To the reader new to type theory, this first section will provide a
gentle introduction to dependently-typed systems. To the reader
familiar with type theory, she will recognise the basis of her
favourite type theory, be it Agda or Coq, for instance. This scene sets,
we have a common ground for discussion.

@@

In the following, we adopt a standard presentation for our type
theory. It is defined by three forms of judgements: context validity
judgements, typing judgements, and equality judgements. Formally,
these judgements are presented as follows:

\[
\begin{array}{ll}
\Gamma \vdash \Valid                & \mbox{\(\Gamma\) is a valid context, giving types to variables} \\
\Gamma \vdash \Bhab{\V{t}}{\V{T}}           & \mbox{term \(\V{t}\) has type \(\V{T}\) in context \(\Gamma\)} \\
\Gamma \vdash \Bhab{\V{s} \equiv \V{t}}{\V{T}}  & \mbox{\(\V{s}\) and \(\V{t}\) are equal at type \(\V{T}\) in context \(\Gamma\)} \\
\end{array}
\]

@@

Each of these judgements is populated by a respective system of
inferences rules. These rules are formulated so as to ensure that the
following implications always hold by induction on derivations:

\[
\begin{array}{l@{\;\Rightarrow\;\;}l}
\Gamma            \vdash \Bhab{\V{t}}{\V{T}}            
    & \Gamma \vdash \Valid\; \wedge\; \Gamma\vdash\Type{\V{T}} \\
\Gamma            \vdash \V{s} \equiv \Bhab{\V{t}}{\V{T}}   
    & \Gamma \vdash \Bhab{\V{s}}{\V{T}} \;\wedge\; \Gamma\vdash \Bhab{\V{t}}{\V{T}} \\
\Gamma ; \xS ; \Delta \vdash \V{J}                      
    & \Gamma \vdash \Bhab{\V{s}}{\V{S}} \;\Rightarrow\; 
          \Gamma ; \Delta[\V{s}/\V{x}] \vdash \V{J}[\V{s}/\V{x}] \\
\end{array}
\]

We are not going to prove the validity of these invariants. They
follow rather straightforwardly from the induction rules. For formal
proofs, we refer the reader to standard presentations of type theory,
such as Luo's seminal work \cite{luo:utt}.

@@

Noticeably, we have adopted a judgemental presentation of
equality. Therefore, we are not tied to a particular implementation
choice. In particular, our system has been modelled in Agda, which
features an intensional equality. On the other hand, it is used in
Epigram, whose equality has a slightly extensional
flavor~\cite{altenkirch:ott}. We expect users of fully extensional
systems to also find their way through this presentation.

@@

The rules ensuring Context Validity 
are unsurprisingly common. They simply guarantee the validity of the
empty context, as well as the linear extension of a context by a
well-typed variable.

\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\Gamma       \vdash \Type{\V{S}}}
     {\Gamma ; \xS \vdash \Valid}\;\V{x}\not\in\Gamma
\]

\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}

The typing judgements (Fig.~\ref{fig:typing-judgements}) are just as
familiar. We recognise the standard presentation of $\Pi$ and $\Sigma$
types, respectively inhabited by lambda terms and dependent
pairs. Naturally, there are rules for function application and
projections of $\Sigma$-types. Equal types can be substituted, thanks
to the conversion rule.

For the sake of presentation, we postulate a $\Set$ in $\Set$
rule. Having this rule makes our type theory inconsistent, by Girard's
paradox~\cite{girard:set-in-set}. However, it has been
shown~\cite{harper:implicit-universe, luo:utt,
  courant:explicit-universe} that a valid stratification can be
inferred, automatically or semi-automatically. In the remaining of our
presentation, we will assume that such a stratification exists, even
though we will keep it implicit. We shall discuss this assumption in
Section~\ref{sec:discussion}.

\input{figure_typing_judgements}

\caption{Typing judgements}
\label{fig:typing-judgements}

@@

Finally, we define the rules governing judgemental equality in
Figure~\ref{fig:judgemental-equality}. We implicitly assume that
judgemental equality respects symmetry, reflexivity, and
transitivity. We capture the computational behavior of the language
through the $\beta$-rules for function application and pairs. Finally,
we implicitly assume that it respects purely syntactic and structural
equality. This includes equality under lambda ($\xi$-rule).

Crucially, being judgemental, this presentation is agnostic in the
notion of equality actually implemented. Indeed, our typing and
equality judgements do not rely on a ``propositional'' equality. This
freedom is a key point in favour of the wide applicability of our
proposal. This judgemental presentation must be read as a
\emph{specification}: our proposal works with any propositional
equality satisfying this specification. Moreover, our lightweight
requirements do not endanger decidability of equality-checking.
Obviously, when implementing our technology in an existing
type-theory, some opportunities arise. We will present some of them
along the course of the paper.

\input{figure_judgemental_equality}

\caption{Judgemental equality}
\label{fig:judgemental-equality}

@@

This completes our presentation of the type theory. Assuming a
stratified discipline of universe, the system we have described enjoy
some very strong meta-theoretical properties. Unlike simply typed
languages, such as Haskell, dependently-typed systems are
\emph{strongly normalising}: every program that type-checks
terminates. Moreover, type-checking is decidable and can therefore be
implemented by a terminating algorithm.
\note{Need some care here. Expansion would be good too. I wanted to
  carry the intuition that we are not the bad guys with a
  non-terminating type-checker.}

@@

\[
\Rule{\Gamma ; \xS ; \Delta \vdash \Valid}
     {\Gamma ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
\]

@@

\[
\Rule{\Gamma       \vdash \Type{\V{S}} \quad
      \Gamma ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
     {\Gamma \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
\]

>>>>

One challenge in writing this paper is to extricate our account of datatypes from what else is new in Epigram 2. In fact, we demand relatively little from the setup, so we shall start with a `vanilla' theory and add just what we need. The reader accustomed to dependent types will recognise the basis of her favourite system; for those less familiar, we try to keep the presentation self-contained.

@@

We adopt a traditional presentation for our type
theory, with three mutually defined systems of judgments:
\emph{context validity}, \emph{typing}, and \emph{equality},
with the following forms:

\[
\begin{array}{ll}
\G \vdash \Valid                & \mbox{\(\G\) is a valid context, giving types to variables} \\
\G \vdash \Bhab{\M{t}}{\M{T}}           & \mbox{term \(\M{t}\) has type \(\M{T}\) in context \(\G\)} \\
\G \vdash \Bhab{\M{s} \equiv \M{t}}{\M{T}}  & \mbox{\(\M{s}\) and \(\M{t}\) are equal at type \(\M{T}\) in context \(\G\)} \\
\end{array}
\]

@@

The rules are formulated to ensure that the
following `sanity checks' hold by induction on derivations
\[
\begin{array}{l@{\;\Rightarrow\;\;}l}
\G            \vdash \Bhab{\M{t}}{\M{T}}            
    & \G \vdash \Valid\; \wedge\; \G\vdash\Type{\M{T}} \\
\G            \vdash s \equiv \Bhab{\M{t}}{\M{T}}   
    & \G \vdash \Bhab{s}{T} \;\wedge\; \G\vdash \Bhab{\M{t}}{\M{T}}
\end{array} \]
and that judgments \(\M{J}\) are preserved by well-typed instantiation.
\[
\G ; \xS ; \Delta \vdash \M{J}            \;\Rightarrow\;      
    \G \vdash \Bhab{\M{s}}{\M{S}} \;\Rightarrow\; 
          \G ; \Delta[\M{s}/\x] \vdash \M{J}[\M{s}/\x] 
\]

%We are not going to prove the validity of these invariants. They
%follow rather straightforwardly from the induction rules. For formal
%proofs, we refer the reader to standard presentations of type theory,
%such as Luo's seminal work \cite{luo:utt}.

@@

We specify equality as a judgment, leaving open the details of its implementation, requiring only a congruence including ordinary computation (\(\beta\)-rules), decided, e.g., by testing \(\alpha\)-equivalence of \(\beta\)-normal forms~\cite{DBLP:journals/jfp/Adams06}. Coquand and Abel feature prominently in a literature of richer equalities, involving \(\eta\)-expansion, proof-irrelevance and other attractions~\cite{DBLP:journals/scp/Coquand96,DBLP:conf/tlca/AbelCP09}. Agda and Epigram 2 support such features, Coq currently does not, but they are surplus to requirements here.

% Therefore, we are not tied to a particular implementation
%choice. In particular, our system has been modelled in Agda, which
%features an intensional equality. On the other hand, it is used in
%Epigram, whose equality has a slightly extensional
%flavor~\cite{altenkirch:ott}. We expect users of fully extensional
%systems to also find their way through this presentation.

@@

Context validity ensures that variables inhabit well-formed
sets.
\[
%% Empty context validity
\Axiom{\vdash \Valid}
\qquad
%% Extend context
\Rule{\G       \vdash \Type{\M{S}}}
     {\G ; \xS \vdash \Valid}\;\x\not\in\G
\]
%
\begin{wstructure}
<- Typing judgements [figure]
    <- Set in Set
        -> For simplicity of presentation
        -> Assume that a valid stratification can be inferred
            <- Harper-Pollack, Luo, Courant
        -> See later discussion
    <- Standard presentation of Pi and Sigma types
\end{wstructure}
%
The basic typing rules
% (Fig.~\ref{fig:typing-judgements})
for tuples and functions are also standard, save that we locally adopt
\(\Set:\Set\), putting presentation before paradox~\cite{girard:set-in-set}.
The usual remedies apply, \emph{stratifying}
\(\Set\)~\cite{harper:implicit-universe, luo:utt,
  courant:explicit-universe}.
\input{figure_typing_judgements}

\paragraph{Notation.} We subscript information needed for type synthesis
but not type checking, e.g., the domain of a \(\LAMBINDER\)-abstraction,
and suppress it informally where clear. Square brackets denote tuples,
with a LISP-like right-nesting convention: \(\sqr{a\;b}\) abbreviates
\(\pair{a}{\pair{b}{\void}{}}{}\).

%We recognise the standard presentation of $\Pi$ and $\Sigma$
%types, respectively inhabited by lambda terms and dependent
%pairs. Naturally, there are rules for function application and
%projections of $\Sigma$-types. Equal types can be substituted, thanks
%to the conversion rule.

%For the sake of presentation, we postulate a $\Set$ in $\Set$
%rule. Having this rule makes our type theory inconsistent, by Girard's
%paradox~\cite{girard:set-in-set}. However, it has been
%shown~\cite{harper:implicit-universe, luo:utt,
%  courant:explicit-universe} that a valid stratification can be
%inferred, automatically or semi-automatically. In the remaining of our
%presentation, we will assume that such a stratification exists, even
%though we will keep it implicit. We shall discuss this assumption in
%Section~\ref{sec:discussion}.

%\begin{figure}
%
%\input{figure_typing_judgements}
%
%\caption{Typing judgements}
%\label{fig:typing-judgements}
%
%\end{figure}

@@

The judgmental equality comprises these computational rules
\input{figure_judgemental_equality} closed under reflexivity,
symmetry, transitivity and structural congruence, even under binders.
We omit the mundane rules which ensure these closure properties for
reasons of space. Given a suitable stratification of \(\Set\), the
computation rules yield a terminating evaluation procedure, ensuring
the decidability of equality and thence type checking.

%Finally, we define the rules governing judgemental equality in
%Figure~\ref{fig:judgemental-equality}. We implicitly assume that
%judgemental equality respects symmetry, reflexivity, and
%transitivity. We capture the computational behavior of the language
%through the $\beta$-rules for function application and pairs. Finally,
%we implicitly assume that it respects purely syntactic and structural
%equality. This includes equality under lambda ($\xi$-rule).

%Crucially, being judgemental, this presentation is agnostic in the
%notion of equality actually implemented. Indeed, our typing and
%equality judgements do not rely on a ``propositional'' equality. This
%freedom is a key point in favour of the wide applicability of our
%proposal. This judgemental presentation must be read as a
%\emph{specification}: our proposal works with any propositional
%equality satisfying this specification. Moreover, our lightweight
%requirements do not endanger decidability of equality-checking.
%Obviously, when implementing our technology in an existing
%type-theory, some opportunities arise. We will present some of them
%along the course of the paper.

%\begin{figure}

%\input{figure_judgemental_equality}
%
%\caption{Judgemental equality}
%\label{fig:judgemental-equality}
%
%\end{figure}

@@

%This completes our presentation of the type theory. Assuming a
%stratified discipline of universe, the system we have described enjoy
%some very strong meta-theoretical properties. Unlike simply typed
%languages, such as Haskell, dependently-typed systems are
%\emph{strongly normalising}: every program that type-checks
%terminates. Moreover, type-checking is decidable and can therefore be
%implemented by a terminating algorithm.
%\note{Need some care here. Expansion would be good too. I wanted to
%  carry the intuition that we are not the bad guys with a
%  non-terminating type-checker.}

@@

\[
\Rule{\G ; \xS ; \Delta \vdash \Valid}
     {\G ; \xS ; \Delta \vdash \Bhab{\V{x}}{\V{S}}}
\]

@@

\[
\Rule{\G       \vdash \Type{\V{S}} \quad
      \G ; \xS \vdash \Bhab{\V{t}}{\V{T}}}
     {\G \vdash \Bhab{\PLAM{\x}{\V{S}} \V{t}}{\PIS{\xS} \V{T}}}
\]
